\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-02-01 jasa/paper-jasa.tex
Section 2.3

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the regression
function
\begin{equation}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_U(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

\subsection{Approximating the reduction map}

In order for this approach to be useful in the likelihood-free setting, we need
to be able to approximate both $s(\mathbf{x})$ and $p(s(\mathbf{x})|\theta)$
based on a finite number of samples $\{\mathbf{x}_i\}$ drawn from the generative
model $p(\mathbf{x}|\theta)$.
Let denote the approximated dimensionality
reduction map $\hat{s}(\mathbf{x})$ and its approximated distribution
$\hat{p}(\hat{s}(\mathbf{x})|\theta)$. \glnote{Explain how to approximate $p(s)$ in practice?}

One strength of this approach is that it factorizes the approximation of the
per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
reproducing the level sets of the per-event likelihood ratio, the density of
$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
the resulting inference will still be valid.\glnote{Shouldn't we elaborate?
Since this is one of our selling points, we should give more details in my
opinion.} \glnote{Some of this is also already outlined in the previous section.}
This point was made by \cite{Neal:2007zz} and is well appreciated by the
particle physics community that typically takes a conservative attitude towards
the use of machine learning classifiers precisely due to concerns about the
calibration $p$-values in the face of nuisance parameters associated to the
simulator.

\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}

\glnote{Add reference to \cite{BiancaZadrozny} here.}

@@

Evaluating the generalized likelihood ratio as defined by
Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
denominator the maximum likelihood estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
\end{equation}
Again, this is made difficult in the likelihood-free setting and it is not
obvious that we can find the same estimators if we are working instead with
$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
the same as the value that maximizes the likelihood ratio with respect to
$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$\glnote{We should mention that the support of $p(\mathbf{x}|\theta_1)$ should cover the support of $p(\mathbf{x}|\theta)$.}.
This allows us to
use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parameterized}
transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$. \glnote{Say that if $s$ is not monotonic with $r$, then it might be bias $\hat\theta$.}

@@

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.

>>>>

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the regression
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_U(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

\subsection{Learning and calibrating $s$}

In order for the proposed approach to be useful in the likelihood-free setting,
we need to be able to approximate both $s(\mathbf{x})$ and
$p(s(\mathbf{x})|\theta)$ based on a finite number of samples $\{\mathbf{x}_i\}$
drawn from the generative model $p(\mathbf{x}|\theta)$.

As outlined above, any consistent probalilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. In particular, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}

Given a reduction map $s$, our results show that a statistic equivalent to the
likelihood ratio can be constructed, provided $p(s(\mathbf{x})|\theta)$ can be
evaluated. Again, we do not have a direct and exact way for evaluating this
density, but an approximation $\hat p(\hat s(\mathbf{x})| \theta)$ can be built
instead, e.g. using density estimation or calibration algorithms, such as
histograms, KDE or isotonic regression, as applied on $\{\hat
s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
model. Most notably, learning such an approximation of $p(s(\mathbf{x})|\theta)$
is a much simpler problem than learning $p(\mathbf{x}|\theta)$, since the
reduction $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the (simpler) informative content of $r(\mathbf{x})$ is preserved.

One strength of the proposed approach is that it factorizes the approximation of the
per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
reproducing the level sets of the per-event likelihood ratio, the density of
$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
the resulting inference will still be valid.
This point was made by \cite{Neal:2007zz} and is well appreciated by the
particle physics community that typically takes a conservative attitude towards
the use of machine learning classifiers precisely due to concerns about the
calibration of $p$-values in the face of nuisance parameters associated to the
simulator.

@@

Evaluating the generalized likelihood ratio as defined by
Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
denominator the maximum likelihood estimator
\begin{equation}\label{eq:mle}
    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
\end{equation}
Again, this is made difficult in the likelihood-free setting and it is not
obvious that we can find the same estimators if we are working instead with
$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
the same as the value that maximizes the likelihood ratio with respect to
$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$ chosen such that the support of $p(\mathbf{x}|\theta_1)$ covers the support of $p(\mathbf{x}|\theta)$.
This allows us to
use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
estimate as
\begin{align}\label{eq:mle_withs}
\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
\end{align}
where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parameterized}
transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
cancels Jacobian factors that vary with $\theta$.

Finally, once the maximum likelihood estimates have been found for both the numerator
and denominator of Eqn.~\ref{eqn:generalized-lr}, the generalized likelihood
ratio can be estimated as outlined in Section~\ref{sec:approx}
for simple hypothesis testing.

@@

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:likelihood-ratio-tests}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.
